title: "Lecture 5 - Bulk RNAseq analysis"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

##Install packages as needed - only run this command once
```{r}
packages_to_install = c(
  "limma", "edgeR", "EDASeq", "RUVSeq", "ffpe", "RColorBrewer", 
  "DESeq2", "org.Mm.eg.db", "NOISeq", "dplyr", "statmod"
)

# Identify missing packages
missing_packages = packages_to_install[!(packages_to_install %in% installed.packages()[,"Package"])]

# Install missing packages
if (length(missing_packages) > 0) {
  install.packages(missing_packages, dependencies = TRUE)
}

# Bioconductor packages
bioc_packages = c("EDASeq", "RUVSeq", "ffpe", "DESeq2", "org.Mm.eg.db", "NOISeq")
missing_bioc = bioc_packages[!(bioc_packages %in% installed.packages()[,"Package"])]

if (length(missing_bioc) > 0) {
  if (!requireNamespace("BiocManager", quietly = TRUE)) install.packages("BiocManager")
  BiocManager::install(missing_bioc)
}
rm(packages_to_install, missing_packages, bioc_packages, missing_bioc)
```

##Load packages - run this command each time this script is loaded
```{r}
require(limma)
require(edgeR)
require(EDASeq)
require(RUVSeq)
require(ffpe)
require(RColorBrewer)
require(DESeq2)
require(org.Mm.eg.db)
require(NOISeq)
require(readxl)
require(dplyr)
require(statmod)
```

##Load sleep deprivation data set 1 (Hor et al. 2019, mouse cortex)
```{r}
# https://www.ncbi.nlm.nih.gov/geo/query/acc.cgi?acc=GSE113754
load("GSE113754_SD1.rda")
SD1 = as.matrix(SD1)
dim(SD1)
head(SD1)
```

##Create a design matrix for the conditions
```{r}
#Create a factor of the conditions for the data matrix
x = as.factor(rep(c("HC","SD5"),times=c(5,5)))
names(x) = colnames(SD1)

###View the factor
x
```
###Set color palette for figures
```{r}
colors = brewer.pal(9, "Set1")
colLib = colors[x]
```

###Remove genes with very low counts
```{r}
countlimit=10
numsamps=3

#Filter the data to only include genes with at least "countlimit" counts in at least "numsamps" samles
filter = rowSums(SD1 > countlimit)
filtered = SD1[which(filter>=numsamps), ]

#See how many genes were in the original and the filtered matrix
nrow(SD1)
nrow(filtered)
```
###Exploratory analysis - PCA
```{r}
###Normalize samples using upper quantile normalization
uq = betweenLaneNormalization(filtered, which = "upper")

###Visualize gene distribution per sample after normalization
plotRLE(uq, col= colLib, outline = FALSE, las = 3, ylim = c(-0.5, 0.5), ylab = "Relative Log Expression", cex.axis = 1, cex.lab = 1)

###Plot samples using PCA
plotPCA(uq, col = colLib, cex = 1, cex.axis = 1, cex.lab = 1, xlim = c(-0.75, 0.75), ylim = c(-0.75, 0.75))
```
###Run standard workflow for bulk DEG analysis
```{r}
###Create design object from design matrix 
design = model.matrix(~x - 1)

###Inspect this matrix
design

###Run standard edgeR workflow 
y = DGEList(counts = filtered, group = x)        ###Create count matrix
y
y = calcNormFactors(y, method = "upperquartile") ###Normalize counts
y
y = estimateDisp(y, design, verbose = TRUE)      ###Estimate dispersion for each gene
y
plot(y$AveLogCPM, y$trended.dispersion)
plot(y$AveLogCPM, y$tagwise.dispersion)
plotBCV(y)
fit = glmQLFit(y, design, robust = TRUE)         ###Calculate F statistics for each gene
plotQLDisp(fit)
qlf1 = glmQLFTest(fit, contrast = c(1, -1))      ###Generate p-values for specific contrast
topUQSD1 = topTags(qlf1, n = Inf)$table          ###Create results table

###Convert ENSEMBL IDs to gene name
genenames=mapIds(org.Mm.eg.db, keys = rownames(topUQSD1), column = 'SYMBOL', keytype = 'ENSEMBL')
topUQSD1$genenames=genenames
topUQSD1$ENSEMBL=rownames(topUQSD1)

head(topUQSD1)

```
###Examine DEG results
```{r}

###Histogram of uncorrected p-values
hist(topUQSD1$PValue, main = "", xlab = "p-value", breaks = 100, ylim = c(0, 1400))


###Check how many significant genes
sum(topUQSD1$FDR < 0.05)

###Create Volcano Plot of all genes, and color DE genes
plot(topUQSD1[, 1], -log10(topUQSD1$FDR), pch = 20, col = "gray", cex= 0.5, ylab = "-log10(FDR adjusted p-value)", xlab= "log2(SD5/HC)", ylim = c(0, 15), xlim = c(-3, 3), cex.lab = 1, cex.axis = 1)
de = rownames(topUQSD1[topUQSD1$FDR <= 0.05, ]) 
points(topUQSD1[de, 1], -log10(topUQSD1[de, "FDR"]), pch = 20, col = colors[2], cex = 1, lwd = 2)

```
###Print up- and down-regulated genes
```{r}
###Sanity check: which condition does positive log fold-change mean?
up1=topUQSD1[which(topUQSD1$logFC>0 & topUQSD1$FDR<0.05),]
head(up1)
down1=topUQSD1[which(topUQSD1$logFC<0 & topUQSD1$FDR<0.05),]
head(down1)

boxplot(10^6*y$counts[up1$ENSEMBL[1],]/colSums(y$counts) ~ y$samples$group,ylab="CPM")
boxplot(10^6*y$counts[down1$ENSEMBL[1],]/colSums(y$counts) ~ y$samples$group,ylab="CPM")

write.table(up1$ENSEMBL,file="Upregulated_in_HC.txt",quote=F,row.names=F)
write.table(down1$ENSEMBL,file="Upregulated_in_SD5.txt",quote=F,row.names=F)
```



#End of workshop part 1#


#Workshop part 2 - Adjusting for hidden confounders


##Load sleep deprivation data sets 1 and 2 (Hor et al. 2019, mouse cortex, Ingiosi et al. 2019)
```{r}
load("GSE113754_SD1.rda")

load("GSE140345_SD2.rda")

#Combine data sets
SD1 = as.matrix(SD1)
SD2 = as.matrix(SD2)
SD = cbind(SD1, SD2)
dim(SD)
head(SD)
```

##Create a design matrix for the conditions
```{r}
#Create a factor of the conditions for the data matrix
x = as.factor(rep(c("HC", "SD5", "HC", "SD6"), c(5,5,5,3)))
names(x) = colnames(SD)

colors = brewer.pal(9, "Set1")
colLib = colors[x]


###View the design matrix and the factor
x
```
###Remove genes with very low counts
```{r}
countlimit=10
numsamps=3

#Filter the data to only include genes with at least "countlimit" counts in at least "numsamps" samles
filter = rowSums(SD > countlimit)
filtered = SD[which(filter>=numsamps), ]

#See how many genes were in the original and the filtered matrix
nrow(SD)
nrow(filtered)
```

###Exploratory analysis - PCA
```{r}
###Normalize samples using upper quantile normalization
uq = betweenLaneNormalization(filtered, which = "upper")

###Visualize gene distribution per sample after normalization
plotRLE(uq, col= colLib, outline = FALSE, las = 3, ylim = c(-0.5, 0.5), ylab = "Relative Log Expression", cex.axis = 1, cex.lab = 1)

###Plot samples using PCA
plotPCA(uq, col = colLib, cex = 1, cex.axis = 1, cex.lab = 1, xlim = c(-0.75, 0.75), ylim = c(-0.75, 0.75))

###What does PC1 separate? A biological or technical condition?

```
###Run DEG expression and check if things make sense
```{r}
design = model.matrix(~x - 1)
y = DGEList(counts = filtered, group = x)
y = calcNormFactors(y, method = "upperquartile")
y = estimateDisp(y, design, verbose = TRUE)
fit = glmQLFit(y, design, robust = TRUE)

# To compare SD6 (3) vs SD5 (2):
qlf1 = glmQLFTest(fit, contrast = c(0, -1, 1))

# To compare SD5 (2) vs HC (1)
qlf2 = glmQLFTest(fit, contrast = c(-1, 1, 0))

# To compare SD6 (3) vs HC (1)
qlf3 = glmQLFTest(fit, contrast = c(-1, 0, 1))

#calculate DE genes for SD6vs SD5, you can do this for every contrast
topUQSD1 = topTags(qlf1, n = Inf)$table
topUQSD2 = topTags(qlf2, n = Inf)$table
topUQSD3 = topTags(qlf3, n = Inf)$table

hist(topUQSD1$PValue, main = "", xlab = "p-value", breaks = 100, ylim = c(0, 1400))
hist(topUQSD2$PValue, main = "", xlab = "p-value", breaks = 100, ylim = c(0, 1400))
hist(topUQSD3$PValue, main = "", xlab = "p-value", breaks = 100, ylim = c(0, 1400))

#check how many DE genes you got
sum(topUQSD1$FDR < 0.05) ###SD5 vs. SD6
sum(topUQSD2$FDR < 0.05) ###Control vs. SD5
sum(topUQSD3$FDR < 0.05) ###Control vs. SD6
```
###Run RUVseq (Remove Unwanted Variation) on the data, using the design and knowledge of control samples
```{r}
#Generate a matrix: each row is a condition, and contains the columns of the data matrix corresponding to that condition
groups = makeGroups(x)
head(groups)


###In RUVseq, k is a paramter representing the number of unwanted factors
s = RUVs(x=uq, cIdx=rownames(uq), scIdx=groups, k= 1)
```

###Now we plot RLE and PCA plots, after normalizing for unwanted factors 
```{r}
plotRLE(s$normalizedCounts, col = colLib, outline = FALSE, las = 3, ylim = c(-0.5, 0.5), ylab= "Relative Log Expression", cex.axis = 1, cex.lab = 1)

plotPCA(s$normalizedCounts, col = colLib, cex = 1, cex.axis = 1, cex.lab = 1, xlim = c(-0.75, 0.75), ylim = c(-0.75, 0.75))
```
###Run DEG analysis, but now including the factor from RUVseq
```{r}
design = model.matrix(~x + s$W - 1)   ###design matrix now includes RUVseq factor s$W
y = DGEList(counts = filtered, group = x)
y = calcNormFactors(y, method = "upperquartile")
y = estimateDisp(y, design, verbose = TRUE)
fit = glmQLFit(y, design, robust = TRUE)

#To compare SD6 (3) vs SD5 (2) at k=1
qlf1_ruv = glmQLFTest(fit, contrast = c(0, -1, 1, 0))

# To compare SD5 (2) vs HC (1) at k=1
qlf2_ruv = glmQLFTest(fit, contrast = c(-1, 1, 0, 0))

# To compare SD6 (3) vs HC (1) at k=1
qlf3_ruv = glmQLFTest(fit, contrast = c(-1, 0, 1, 0))

#calculate DE genes for SD6vs SD5, you can do this for every contrast
topUQSD1 = topTags(qlf1_ruv, n = Inf)$table
topUQSD2 = topTags(qlf2_ruv, n = Inf)$table
topUQSD3 = topTags(qlf3_ruv, n = Inf)$table

hist(topUQSD1$PValue, main = "", xlab = "p-value", breaks = 100, ylim = c(0, 1400))
hist(topUQSD2$PValue, main = "", xlab = "p-value", breaks = 100, ylim = c(0, 1400))
hist(topUQSD3$PValue, main = "", xlab = "p-value", breaks = 100, ylim = c(0, 1400))

#check how many DE genes you got
sum(topUQSD1$FDR < 0.05) ###SD5 vs. SD6
sum(topUQSD2$FDR < 0.05) ###Control vs. SD5
sum(topUQSD3$FDR < 0.05) ###Control vs. SD6
```
#Workshop part 3 - Running DAVID (interactive web interface)
https://david.ncifcrf.gov/summary.jsp



